---
title: "Starting with with the dplyr, ggplot2 packages, and the pipe operator |>"
author: "Han Olff"
date: "2024-08-29"
format: html
editor: 
  markdown: 
    wrap: 72
editor_options:
  chunk_output_type: console
---

# Working with the dplyr and ggplot2 packages

Load the required libraries :

```{r message = FALSE, warning = FALSE}
# restore your library
renv::restore()
# clear the R environment
rm(list = ls())

# load the required packages
library(tidyverse)  # this loads a group of related packages, including readr, dplyr and ggplot2
```

We will work with an online database of the Schiermonnikoog transect
study, that you will collect additional data for next week in the field.
In this example, we work with measurements of cockles (a bivalve
mollusc) on their width and length. From the database Epibenthos, we
will work with the table
[FactCockles](https://docs.google.com/spreadsheets/d/1E1vUWAHhse7fhjBf94Kiog3Rsj7IkMuFtP-BTQX2oI8/edit?gid=1538766002#gid=1538766002).
See the documentation of the different variables in the table
[MetVariables](https://docs.google.com/spreadsheets/d/1E1vUWAHhse7fhjBf94Kiog3Rsj7IkMuFtP-BTQX2oI8/edit?gid=1290622213#gid=1290622213)

### Read the datafile from the Google sheets database

To read the data in R, you first need to know the published csv link of
the FactCockles table. In the database, you can find this link in the
table MetTables. It is produced in Google Sheets throught the menu
File/Share/Publish to web and then selecting the table and output as
csv.

Read the FactCockleSize table with read_csv, which reads it as tibble
(formatted dataframe)

```{r}
FactCockleSize <- readr::read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSpormjGiM4uWMTIZYm9Upt5j1Ige_Wu9CrhGGdvjXwxmHP2R6S7JDyEmiGTn3fXihl5s5yBn_sdo8h/pub?gid=1538766002&single=true&output=csv")
print(FactCockleSize)
nrow(FactCockleSize)
```

Plot with ggplot the relation between cockle thickness (thickness_mm, as
x) and length (length_mm), showing each year with a different color, and
add a regression line through all the years

```{r}
#delete big outlier
FactCockleSize |> dplyr::filter(thickness_mm<500)

FactCockleSize <- readr::read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSpormjGiM4uWMTIZYm9Upt5j1Ige_Wu9CrhGGdvjXwxmHP2R6S7JDyEmiGTn3fXihl5s5yBn_sdo8h/pub?gid=1538766002&single=true&output=csv") |> dplyr::filter(CockleObs_ID!=1531, CockleObs_ID!=469) |> dplyr::mutate(year=factor(year))

ggplot2::ggplot(data=FactCockleSize,
                mapping=aes(x=thickness_mm,y=length_mm,color=year))+
  geom_point()+
  geom_smooth(method="lm") + 
  theme_light()
#find the other outlier
FactCockleSize |> dplyr::filter(thickness_mm>10 & length_mm<5)

#also remove cockle 469

```

Further explore the plot with a regression line through all the data

```{r}
ggplot2::ggplot(data=FactCockleSize,
                mapping=aes(x=thickness_mm,y=length_mm))+
  geom_point(mapping=aes(color=year))+
  geom_smooth(method="lm",color="yellow") + 
  theme_dark()

#calculate a linear regression
model_lm <- lm(thickness_mm~length_mm, data=FactCockleSize)
summary(model_lm)
#equation: thickness_mm=-0.353+0.669*length_mm
# a cockle of length 20 mm has an predicted width of:
-0.353+0.669*20
#this regression is significant, R2=0.90, P<0.001
```

Make same plot but showing a separate regression line per year

```{r}
# color the points by year, but plot one regression line
```

Make a panel plot where with each year is shown as a separate graph

```{r}
drop_na(FactCockleSize) |>
ggplot2::ggplot(mapping=aes(y=thickness_mm,x=length_mm))+
  geom_point(mapping=aes(color=year))+
  geom_smooth(method="lm",color="yellow") +
  facet_wrap(~year)

```

We conclude from this that: \* there were two important outliers in the
dataset that were removed after visual inspection \* the regression
between length and width is abou the same for every year, we safely can
use only length as a proxy for the biomass of an individual cockle
